Comparing Parallel Sorting Algorithms on a High Performance Computing Cluster Using OpenMP

### 2020-11-16 paper/Paper.tex
cite added
Quicksort implementation description
Added description of quicksort, not sure of the formatting

    \subsection{Mergesort}
    Mergesort is a divide-and-conquer sorting algorithm that recursively divides an array into subarrays and merges them together in such a way that the original array is sorted. 
    The operation of this procedure on an array of eight elements is depicted in Figure \ref{mrg}.  
    \begin{figure}[h]
        \includegraphics[width=6cm]{merge.png} 
        \caption{Mergesort diagram from \cite{cormen_introduction_2009}}
        \label{mrg}
    \end{figure}
    Mergesort has a time complexity of $O(nlog(n))$ in both the average and worst case, although constant factors can make it worse than Quicksort in practice. 
    Given the divide-and-conquer paradigm that Mergesort uses, it is a natural candidate for parallelization. Mergesort follows this paradigm as follows: First the input array (of size $n$) is recursively divided in two until there are $n$ single-element subarrays (divide). Each single-element array is trivially sorted (conquer). From there, the sorted subarrays are merged together, resulting in a single array that is equivalent to the input array but in sorted order (combine). Given the mechanics of the algorithm, there are two primary ways to parallelize Mergesort: (1) Parallelize the divide stage (i.e., the two recursive calls to the Mergesort procedure) and (2) Parallelize the combine stage (i.e., the Merge procedure). It is possible to implement either (1), (2), or both. According to \cite{cormen_introduction_2009}, only parallelizing the recursive calls to Mergesort will result in diminishing returns on performance as the number of processors grows to more than a few dozen. In order to efficiently scale up to hundreds of processors, parallelizing the Merge procedure is necessary because it becomes the performance bottleneck.
    Our implementation of parallelized Mergesort is significantly based on [SOURCE]. The implementation parallelizes the recursive calls to Mergesort while implementing a serial Merge procedure. There are a couple of reasons for this design decision. First, the HPC architecture we are using (the Case Western Reserve Markov cluster) has well under one hundred CPU cores per node. Thus, parallel Mergesort with only the divide stage parallelized theoretically scales well within this core-count range. Second, parallelizing the Merge procedure proved to be difficult with OpenMP due to a lack of low-level control over threads.

@@

    \subsection{Quicksort}
    The final sort discussed here is Quicksort, another $O(n log(n))$ comparison sort. 
    Developed in 1961 \cite{hoare_algorithm_1961}, Quicksort is a partitioning sort that works by selecting a pivot element in an array and partitioning based on a comparison to the pivot, as shown in Figure \ref{qck}. 
    \begin{figure}[h]
        \includegraphics[width=6cm]{Quicksort.png} 
        \caption{Quicksort diagram. \href{https://www.techiedelight.com/quicksort/}{Source}}
        \label{qck}
    \end{figure}
    In particular, Quicksort is perhaps the most parallelizable sorting algorithm, and can achieve a linear speedup with few modifications. \cite{blelloch_programming_1996}
    This is because each subarray can be sorted independently, and this leads to speedup. 
    % TODO @Jason add detail on your implementation
    \section{Methodology}
	For our project, we used...
    For our project, we implemented our sorting algorithms in the C programming language. OpenMP \cite{openmp08} is used for achieving parallelization for our implementations of parallelized Mergesort and Quicksort.
    OpenMP facilitates multithreaded shared memory parallelism through the use of \textit{\#pragma} preprocessor directives in our C code.
	We ran our algorithms on Case Western Reserve University's HPC Markov cluster, which uses Intel Xeon x86\_64 Processors.

>>>>

    \subsection{Mergesort}
    Mergesort is a divide-and-conquer sorting algorithm that recursively divides an array into subarrays and merges them together in such a way that the original array is sorted. 
    The operation of this procedure on an array of eight elements is depicted in Figure \ref{mrg}.  
    \begin{figure}[h]
        \includegraphics[width=6cm]{merge.png} 
        \caption{Mergesort diagram from \cite{cormen_introduction_2009}}
        \label{mrg}
    \end{figure}
    Mergesort has a time complexity of $O(nlog(n))$ in both the average and worst case, although constant factors can make it worse than Quicksort in practice. 
    Given the divide-and-conquer paradigm that Mergesort uses, it is a natural candidate for parallelization. Mergesort follows this paradigm as follows: First the input array (of size $n$) is recursively divided in two until there are $n$ single-element subarrays (divide). Each single-element array is trivially sorted (conquer). From there, the sorted subarrays are merged together, resulting in a single array that is equivalent to the input array but in sorted order (combine). Given the mechanics of the algorithm, there are two primary ways to parallelize Mergesort: (1) Parallelize the divide stage (i.e., the two recursive calls to the Mergesort procedure) and (2) Parallelize the combine stage (i.e., the Merge procedure). It is possible to implement either (1), (2), or both. According to \cite{cormen_introduction_2009}, only parallelizing the recursive calls to Mergesort will result in diminishing returns on performance as the number of processors grows to more than a few dozen. In order to efficiently scale up to hundreds of processors, parallelizing the Merge procedure is necessary because it becomes the performance bottleneck.
    Our implementation of parallelized Mergesort is significantly based on \cite{radenski_mergesort}. The implementation parallelizes the recursive calls to Mergesort while implementing a serial Merge procedure. There are a couple of reasons for this design decision. First, the HPC architecture we are using (the Case Western Reserve Markov cluster) has well under one hundred CPU cores per node. Thus, parallel Mergesort with only the divide stage parallelized theoretically scales well within this core-count range. Second, parallelizing the Merge procedure proved to be difficult with OpenMP due to a lack of low-level control over threads.

@@

    \subsection{Quicksort}
    The final sort discussed here is Quicksort, another $O(n log(n))$ comparison sort. 
    Developed in 1961 \cite{hoare_algorithm_1961}, Quicksort is a partitioning sort that works by selecting a pivot element in an array and partitioning based on a comparison to the pivot, as shown in Figure \ref{qck}. 
    \begin{figure}[h]
        \includegraphics[width=6cm]{Quicksort.png} 
        \caption{Quicksort diagram. \href{https://www.techiedelight.com/quicksort/}{Source}}
        \label{qck}
    \end{figure}
    In particular, Quicksort is perhaps the most parallelizable sorting algorithm, and can achieve a linear speedup with few modifications. \cite{blelloch_programming_1996}
    This is because each subarray can be sorted independently, and this leads to speedup.
    The fundamental steps of quicksort is to partition the array around a selected “pivot” point to produce two subarrays. These subarrays are then recursively sorted until the entire array of elements are in the correct order. Intuitively, one can see how to parallelize based on this fundamental divide-and-conquer algorithm. In OpenMP, the “task” directive can tell OpenMP to dedicate a thread to a specific block of code labeled by the task. In this case, when the two subarrays are selected, each recursive call to the quicksort method gets a dedicated OpenMP task initiated to create a parallel region of threads to be called later. This increases efficiency incredibly when allocating more and more OpenMP threads and correlating core count for quicksort to use. Our graphical analysis below indicates this speedup, along with the dependence on an increasing number of elements to sort.
    \section{Methodology}
	For our project, we used...
    For our project, we implemented our sorting algorithms in the C programming language. OpenMP \cite{openmp08} is used for achieving parallelization for our implementations of parallelized Mergesort and Quicksort.
    OpenMP facilitates multithreaded shared memory parallelism through the use of \textit{\#pragma} preprocessor directives in our C code.
	We ran our algorithms on Case Western Reserve University's HPC Markov cluster, which uses Intel Xeon x86\_64 Processors.
